---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r}

dfActivity <- read.table(unz("activity.zip", "activity.csv"), header=TRUE, sep=',')
print(head(dfActivity))


```

## What is mean total number of steps taken per day?

```{r}

dfTotalbyDay = setNames(aggregate(dfActivity$steps, by=list(dfActivity$date), FUN=sum),
                        c("date", "totalSteps")
                        )


print(dfTotalbyDay)
sprintf("Total steps: %d, Number days not NA: %d", sum(dfTotalbyDay$totalSteps, na.rm = TRUE), 
                                                  sum(!is.na(dfTotalbyDay$totalSteps)))

h1 <- hist(dfTotalbyDay$totalSteps,
                        breaks = 10, 
                        main='Histogram Total Steps per Day', 
                        xlim=c(0,25000),
                        ylim=c(0,20),
                        col='cornsilk', 
                        xlab='Steps', 
                        ylab='Frequency'
                   )

text(h1$mids,h1$counts,labels=h1$counts, adj=c(0.5, -0.5))


```


```{r}
mean_totalsteps = mean(dfTotalbyDay$totalSteps, na.rm = TRUE )
median_totalsteps = median(dfTotalbyDay$totalSteps, na.rm = TRUE )

sprintf("Mean total steps: %s", format(mean_totalsteps,big.mark=",",scientific=FALSE))
sprintf("Median total steps: %s", format(median_totalsteps,big.mark=",",scientific=FALSE))
```




## What is the average daily activity pattern?

```{r}
dfDailyActivity = aggregate(dfActivity$steps, 
                            by=list(dfActivity$interval), 
                            FUN=mean,
                            na.rm=TRUE)

plot(dfDailyActivity, type='l')
                            
```


Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
interval = dfDailyActivity[ which.max(dfDailyActivity$x), 1]

sprintf("5 min interval with maxium steps: %d", interval)

```

## Imputing missing values


Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
count = sum(is.na(dfActivity$steps))

sprintf("Number of missing step values: %d", count)
```


Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}

# Use average daily steps in an interval (across all days) as a replacement 
# for NA
dfReplacement <- setNames( aggregate(dfActivity$steps, 
                                     by=list(dfActivity$interval), 
                                     FUN=mean,
                                     na.rm=TRUE),
                           c("interval", "aveSteps"))
print(dfReplacement)


# Create a new dataframe and replace NA values
dfActivityNoNA <- dfActivity
dfActivityNoNA$oldValue <- dfActivityNoNA$steps

sprintf("Number of NA Before:  %d", sum(is.na(dfActivityNoNA$steps)))

for(i in 1:nrow(dfActivityNoNA)) {       # for-loop over rows
     if (is.na(dfActivityNoNA[i, 1])) {  # if step value for row i is NA
          # Replacement value is the aveSteps in dfReplacement 
          # for the interval value found in DfActivityNoNA
          replacement_value <- dfReplacement[dfReplacement$interval == dfActivityNoNA[i,3], 2]
          dfActivityNoNA[i,1] <- replacement_value
          #dfActivityNoNA[i,1] <- 0
          replacement_value     }
}
print(dfActivityNoNA)
sprintf("Number of NA After:  %d", sum(is.na(dfActivityNoNA$steps)))

```

Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?


```{r}

dfTotalbyDayNoNA = setNames(aggregate(dfActivityNoNA$steps, by=list(dfActivityNoNA$date), FUN=sum),
                            c("date", "totalSteps")
                            )
print(dfTotalbyDayNoNA)
sprintf("Total steps: %f.2, Number days not NA: %d", sum(dfTotalbyDayNoNA$totalSteps), 
                                                   sum(!is.na(dfTotalbyDayNoNA$totalSteps)))



h2 <- hist(dfTotalbyDayNoNA$totalSteps,
                            breaks = 10, 
                            main='Total Steps per Day with Imputed Missing Values', 
                            xlim=c(0,25000),
                            ylim=c(0,25),
                            col='darkseagreen3', 
                            xlab='Step', 
                            ylab='Frequency'
                     )

text(h2$mids,h2$counts,labels=h2$counts, adj=c(0.5, -0.5))


mean_totalstepsNoNA = mean(dfTotalbyDayNoNA$totalSteps)
median_totalstepsNoNA = median(dfTotalbyDayNoNA$totalSteps)

sprintf("Mean total steps (with imputed missing values): %s", 
            format(mean_totalstepsNoNA,big.mark=",",scientific=FALSE))
sprintf("Median total steps (with imputed missing values): %s", 
            format(median_totalstepsNoNA,big.mark=",",scientific=FALSE))
```

## Are there differences in activity patterns between weekdays and weekends?


Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
Make a panel plot containing a time series plot (i.e. \color{red}{\verb|type = "l"|}type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.


```{r}

dfActivityNoNA$asDate <- as.Date(dfActivityNoNA$date)
dfActivityNoNA$day <- "weekday"
for(i in 1:nrow(dfActivityNoNA)) {       # for-loop over rows
     if (grepl("S(at|un)", weekdays(dfActivityNoNA[i, 5]))) {
         dfActivityNoNA[i, 'day'] <- "weekend"
     }
}
dfActivityNoNA$day <- as.factor(dfActivityNoNA$day)

summary(dfActivityNoNA)


dfDayActivity = setNames(aggregate(dfActivityNoNA$steps, 
                              by=list(dfActivityNoNA$interval, dfActivityNoNA$day), 
                              FUN=mean,
                              na.rm=TRUE),
                         c("interval", "day", "aveSteps"))

par(mfrow=c(1,2))        

df1 = X$weekend[ ,c(1,3)]
df2 = X$weekday[ ,c(1,3)]

plot(df1, type = "l")  # Plot with Base R
plot(df2, type = "l")  # Plot with Base R

#plot(dfDayActivity$Group.1, dfDailyActivity$x, type='l')

```




